---
title: "codes"
output: html_document
---

```{r, message=FALSE, warning=FALSE, echo=FALSE}
# Load packages
library(MASS)
library(caret)
library(tidyverse)
library(survival)
library(splines) 
library(broom)
```

```{r}
# Import data
breast_cancer <- read_csv('data/Project_2_data.csv', show_col_types = FALSE) %>%  
  janitor::clean_names() %>% 
  mutate(
    race = as_factor(race),
    marital_status = as_factor(marital_status),
    x6th_stage = as_factor(x6th_stage),
    differentiate = as_factor(differentiate),
    a_stage = as_factor(a_stage),
    estrogen_status = as_factor(estrogen_status),
    progesterone_status = as_factor(progesterone_status), 
    # Calculates the proportion of positive regional nodes examined
    regional_prop = reginol_node_positive / regional_node_examined, 
    # Transform tumor_size
    tumor_size = log(tumor_size), 
    x6th_stage = fct_collapse(x6th_stage, "IIIB_IIIC" = c("IIIB", "IIIC")), 
  ) %>% 
  dplyr::select(-c(t_stage, n_stage, differentiate, reginol_node_positive, regional_node_examined, survival_months))

# Convert Status to binary variable (Dead=1, Alive=0)
breast_cancer$status <- ifelse(breast_cancer$status == "Dead", 1, 0)

breast_cancer <- breast_cancer %>%
  mutate(regional_prop_spline = ns(regional_prop, df = 3)) 
```

Build the full model 

```{r}
full_model <- glm(status ~ age + race + marital_status + x6th_stage + grade + 
                  estrogen_status * progesterone_status + tumor_size + regional_prop, 
                  data = breast_cancer, family = binomial())
full_model <- glm(
  status ~ age + race + marital_status + x6th_stage + grade + 
           estrogen_status * progesterone_status + tumor_size + 
           ns(regional_prop, df = 3), 
  data = breast_cancer,
  family = binomial
)
```

Use AIC to optimize the model  

```{r}
opt_model <- stepAIC(full_model, direction = "both", trace = FALSE)
summary(opt_model)
car::vif(opt_model)
```


```{r}
breast_cancer$status <- as.factor(breast_cancer$status)
# 10-fold cross validation
set.seed(123)
train_control <- trainControl(method = "cv", number = 10)
cv_model <- train(
  status ~ age + race + marital_status + x6th_stage + grade +
           estrogen_status * progesterone_status + tumor_size + regional_prop,
  data = breast_cancer,
  method = "glm",
  family = binomial,
  trControl = train_control
)

print(cv_model)
```

```{r}
breast_cancer$pred_prob <- predict(opt_model, type = "response")

library(pROC)
roc_curve <- roc(breast_cancer$status, breast_cancer$pred_prob)
plot(roc_curve, main = "ROC Curve")
auc(roc_curve)

breast_cancer$status <- as.numeric(as.character(breast_cancer$status))
breast_cancer$pred_prob <- predict(opt_model, type = "response")
brier_score <- mean((breast_cancer$pred_prob - breast_cancer$status)^2)
print(paste("Brier Score:", round(brier_score, 4)))
```

Summary: 

**status∼age + race + marital_status + x6th_stage + grade + estrogen_status + progesterone_status + tumor_size + ns(regional_prop, df = 3)**

Key Features in the Model:
1. Linear Predictors:
`age`: Age of the patient.
`race`: Ethnicity group (e.g., White, Black, Other).
`marital_status`: Marital status of the patient.
`x6th_stage`: Tumor staging category.
`grade`: Cancer grade (including Grade IV as Anaplastic).
`tumor_size`: Transformed tumor size (log-transformed).
2. Nonlinear Relationship:
`regional_prop`: Modeled using natural splines (ns) with 3 degrees of freedom to capture its non-linear relationship.
3. Interaction Term:
estrogen_status * progesterone_status: To account for the interaction effect between estrogen and progesterone statuses.
4. Model Performance:
AUC: 0.7513 — indicating good discriminatory power.
Brier Score: 0.1119 — reflecting well-calibrated probability predictions.
Accuracy: 85.5% (with 10-fold cross-validation).
Kappa: 0.175, showing moderate agreement.

Why This Model?
Balance of Performance and Interpretability:
A Generalized Linear Model (GLM) with natural splines and interaction terms achieves a good trade-off between predictive accuracy and interpretability.
Handling Non-Linearity:
By applying splines to regional_prop, the model effectively captures its nonlinear impact on the outcome.

```{r}
breast_cancer$pred_prob <- predict(opt_model, type = "response")

# Create a new column for simplified race groups
breast_cancer <- breast_cancer %>%
  mutate(race_group = case_when(
    race == "White" ~ "White",
    race == "Black" ~ "Black",
    TRUE ~ "Other"
  ))

# Function to evaluate performance metrics
evaluate_performance <- function(data, group_name) {
  cat("\nPerformance for:", group_name, "\n")
  
  # Calculate ROC and AUC
  roc_curve <- roc(data$status, data$pred_prob)
  auc_value <- auc(roc_curve)
  cat("AUC:", auc_value, "\n")
  
  # Binary predictions for accuracy
  data$pred_class <- ifelse(data$pred_prob > 0.5, 1, 0)
  confusion <- confusionMatrix(as.factor(data$pred_class), as.factor(data$status), positive = "1")
  
  cat("Accuracy:", confusion$overall['Accuracy'], "\n")
  cat("Kappa:", confusion$overall['Kappa'], "\n")
}

# White group
white_group <- breast_cancer %>% filter(race_group == "White")
evaluate_performance(white_group, "White")

# Black group
black_group <- breast_cancer %>% filter(race_group == "Black")
evaluate_performance(black_group, "Black")

# Other group
other_group <- breast_cancer %>% filter(race_group == "Other")
evaluate_performance(other_group, "Other")
```

Conclusion:  
The White group achieves the best balance between AUC and Accuracy.  
The Black group has acceptable accuracy but a lower AUC, suggesting room for improvement.  
The Other group has high accuracy but low AUC and Kappa, indicating overfitting or poor calibration.  






